Linear instability underlying synchronous waving of marine grass in flow

### 2014-09-21 Paper/grass.tex
more update with MAHA comments and figures

To understand the onset of, we first calculate the steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ as a solution of \eqref{navier-stokes}, and examine its linear stability. 
Such a steady state flow driven by constant pressure gradient satisfies 
\begin{equation}
 -{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0
\end{equation}
where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< 2H$. In the limit of large grass density, shear stress exerted by the bottom surface is estimated to be negligible in comparison with the drag and hence the bottom surface ($y=0$) is approximated by zero shear.   
%The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=2H$) of the fluid layer, 
%as can be seen from the comparison between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{basicflow}. 
A comparison of laboratory mean profile with the steady flow profile is shown in figure \ref{basicflow}.
Flow within grass has approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from the balance between drag and with pressure gradient. Flow outside has a simple parabolic velocity profile due to the balance between viscous forces and the pressure gradient. 
Continuity of flow speed and shear stresses at canopy top gives rise to a boundary layer 
%in the flow profile 
near the tip of the grass.

This boundary layer near the canopy top results from purely local dynamics independent of the influence of the boundaries, and we identify it as a free shear layer\cite{Ghisal02} in the previous explanation of \monami. 
In our model shear layer thickness can be estimated from the the balance between viscous forces and drag in the boundary layer near canopy top along with the continuity of shear stress at the grass tip. This leads to a shear layer thickness $\delta = (\Rey\Ndg)^{-1/3} H$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow, and $U_0 = {(dP/dx)~H^2}/{\mu}$ is scale for velocity. 
This dependence of the boundary layer thickness on the grass density $N_g$, shown in Figure \ref{basicflow}, gives us a way to systematically control it and investigate its effect on the instability mechanism.
The asymptotic regime of a thin boundary layer is expected to hold for $\Rey \Ndg \gtrsim 100$; minor departures from this scaling can be seen in the inset of figure \ref{basicflow}.

\begin{figure*}
\begin{tabular}{cccc}
{\includegraphics[scale = 0.67]{Set4_dens28_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens30_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens32_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens34_imgsc}} \\
{\includegraphics[scale = 0.67]{Set5_dens38_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens40_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens42_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens46_imgsc}} \\
{\includegraphics[scale = 0.67]{Set6_dens32_imgsc}} &
{\includegraphics[scale = 0.67]{Set6_dens34_imgsc}} &
{\includegraphics[scale = 0.67]{Set6_dens36_imgsc}} &
{\includegraphics[scale = 0.67]{Set6_dens38_imgsc}}
\end{tabular}
\caption{$Re(\sigma)$ as function of wavenumber and Reynolds number for a specified grass number density $\Ndg$ along with the neutral curve ($Re(\sigma)$=0), for parameters shown in the corresponding panel.  
As $\Ndg \propto N_g$ increases, the unstable region splits into two labeled as ``Mode 1'' and ``Mode 2''. 
For $\Ndg$ below a critical value, the Mode 1 sets the threshold $\Rey$ whereas above the critical $\Ndg$ onset is determined by Mode 2.}
\label{K_Re_sigma_set3}
\end{figure*}

Considering small perturbations $u, v, p$ about the base profile $U$ and $P$ respectively, equation \eqref{navier-stokes} expanded to linear order in the perturbations yields
\begin{equation}
\begin{split}
\rho(u_t+U u_x+vU_y) &= -p_x+ {\mu}\nabla^2u-2S\rho C_{N}dN_{g}Uu, \\
\rho(v_t+ Uv_x) &= -p_y+ {\mu}\nabla^2v, \hspace{0.3cm} \nabla\cdot\bu=0.
\end{split}
\end{equation}
These equations can be non-dimensionalized using half channel height $H$, velocity $U_0$, and the associated advection time $H/U_0$. 
We are left with three non-dimensional parameters in the model, viz. $\Rey$, $\Ndg$, and the submergence ratio of the grass $h_g/H$. 
We also use $\delta/H$ in lieu of $\Ndg$ to parametrize the grass density and help elucidate the mechanism of the instability. 
With these scaling, and using a stream function $\psi$ with $u = \psi_{y}, v= -\psi_x$ to satisfy mass balance, we seek a wave solution of 
the form $\left(u,v,\psi \right)= \left(\hat u, \hat v, \hat\phi \right)e^{ikx+\sigma t}$ to  obtain a modified Orr-Sommerfield equation. 
The eigenvalue $\sigma$ for a given $k$ provides growth rate and frequency associated with a perturbation of wavelength $2\pi/k$ through the solution of
\begin{equation}
\begin{split}
\left(D^2 -k^{2} \right)^2\phi &= \Rey \left[ \left({\sigma}+ikU\right) \left(D^2-k^2\right) -ikU_{yy}\right]\phi \\
&+D\left(2R \Ndg S U D \phi\right),
\label{Orr-somerfield}
\end{split}
\end{equation}
where $D=d/dy$, and subject to the boundary conditions $D\phi = D^2\phi = 0$ at $y=0$ and $y=2$.

Comparison of frequency (Im$(\sigma)$) of the fastest growing mode with that of experimental observations of frequencies in
the lab scale experiments, for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field, 
is shown in Figure \ref{Re_vs_delta}. 
The observed frequencies are associated with the peaks in velocity spectra, frequency of $\monami$ and frequency of 
vortex passage \cite{Ghisal02}, and compare well with the predicted frequencies.

A threshold in $\Rey$ of O(100), above which the flow is unstable (Re($\sigma$)$>$0), naturally emerges from the 
solution of \eqref{Orr-somerfield}. 
The dependence of this critical $\Rey$, and the corresponding fastest growing wavenumber $k$, on $\delta/H$ and $\hg/H$ is shown 
in figure \ref{Re_vs_delta}, and is found to compare well with experimental observations\cite{Ghisal02}.
The threshold Reynolds number increases with the grass density, indicating a competition between the 
destabilizing shear in the flow, and the stabilizing effect of damping due to vegetation drag.
Previous linear stability calculations undertaken for terrestrial grass do not exhibit a threshold flow condition 
either due to the exclusion of the vegetation drag in their model\cite{Raupach96}, or because of the over-simplifying 
assumptions such as a tanh- or piece-wise linear mean velocity profile\cite{Raupach96,Delangre06}.
Regardless of the assumptions, the introduction of vegetation drag is observed to damp the linear modes\cite{Delangre06}.

>>>>

To understand the onset of, we first calculate the steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ as a solution of \eqref{navier-stokes}, and examine its linear stability. 
Such a steady state flow driven by constant pressure gradient satisfies 
\begin{equation}
 -{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0
\label{base_equ}
\end{equation}
where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< 2H$. In the limit of large grass density, shear stress exerted by the bottom surface is estimated to be negligible in comparison with the drag and hence the bottom surface ($y=0$) is approximated by zero shear.   
%The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=2H$) of the fluid layer, 
%as can be seen from the comparison between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{basicflow}. 
A comparison of laboratory mean profile with the steady flow profile is shown in figure \eqref{basicflow}.
Flow within grass has approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from the balance between drag and with pressure gradient. Flow outside has a simple parabolic velocity profile due to the balance between viscous forces and the pressure gradient. 
Continuity of flow speed and shear stresses at canopy top gives rise to a boundary layer 
%in the flow profile 
near the tip of the grass.

This boundary layer near the canopy top results from purely local dynamics independent of the influence of the boundaries, and we identify it as a free shear layer\cite{Ghisal02} in the previous explanation of \monami. 
In our model the balance between viscous forces and the drag force in the shear layer 
%$C_NdN_g\delta^2u/\nu \sim O(1)$ 
together with the continuity of shear stress across the grass tip 
%$u \sim \delta U_o/H$
 leads to a shear layer thickness $\delta=(\Rey\tilde{N_g})^{-1/3}H$.
% and velocity $u \sim \delta U_0$ in the shear layer. 
Where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow, and $U_0 = {(dP/dx)~H^2}/{\mu}$ is scale for velocity. 
This dependence of the boundary layer thickness on the grass density $N_g$, shown in Figure \ref{basicflow}, gives us a way to systematically control it and investigate its effect on the instability mechanism.
The asymptotic regime of a thin boundary layer is expected to hold for $\Rey \Ndg \gtrsim 100$; minor departures from this scaling can be seen in the inset of figure \ref{basicflow}.

\begin{figure*}
\begin{tabular}{cccc}
{\includegraphics[scale = 0.67]{Set4_dens28_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens30_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens32_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens34_imgsc}} \\
{\includegraphics[scale = 0.67]{Set5_dens38_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens40_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens42_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens46_imgsc}} \\
%{\includegraphics[scale = 0.67]{Set6_dens32_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens34_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens36_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens38_imgsc}}
\end{tabular}
\caption{$Re(\sigma)$ as function of wavenumber and Reynolds number for a specified grass number density $\Ndg$ along with the neutral curve ($Re(\sigma)$=0), for parameters shown in the corresponding panel.  
As $\Ndg \propto N_g$ increases, the unstable region splits into two labeled as ``Mode 1'' and ``Mode 2''. 
For $\Ndg$ below a critical value, the Mode 1 sets the threshold $\Rey$ whereas above the critical $\Ndg$ onset is determined by Mode 2.}
\label{K_Re_sigma_set3}
\end{figure*}

Considering small perturbations $u, v, p$ about the base profile $U$ and $P$ given by the solution of \eqref{base_equ}. Substituting them in equation \eqref{navier-stokes} and expanding  to linear order in the perturbation variable we obtain.
\begin{equation}
\begin{split}
\rho(u_t+U u_x+vU_y) &= -p_x+ {\mu}\nabla^2u-2S\rho C_{N}dN_{g}Uu, \\
\rho(v_t+ Uv_x) &= -p_y+ {\mu}\nabla^2v, \hspace{0.3cm} \nabla\cdot\bu=0.
\end{split}
\end{equation}
These equations can be non-dimensionalized using half channel height $H$, velocity $U_0$, and the associated advection time $H/U_0$. 
We are left with three non-dimensional parameters in the model, ie. $\Rey$, $\Ndg$, and the submergence ratio of the grass $h_g/H$. 
We also use $\delta/H$ in lieu of $\Ndg$ to parametrize the grass density and help elucidate the mechanism of the instability. 
With these scaling, and using a stream function $\psi$ with $u = \psi_{y}, v= -\psi_x$ to satisfy mass balance, we seek a wave solution of 
the form $\left(u,v,\psi \right)= \left(\hat u(y), \hat v(y), \hat\phi(y) \right)e^{ikx+\sigma t}$ to  obtain a modified Orr-Sommerfield equation. 
The growth rate $\sigma$ for a given wave number $k=2\pi /\lambda$ can be shown to satisfy the eigenvalue problem.
% provides growth rate and frequency associated with a perturbation of wavelength $2\pi/k$ through the solution of
\begin{equation}
\begin{split}
\left(D^2 -k^{2} \right)^2\phi &= \Rey \left[ \left({\sigma}+ikU\right) \left(D^2-k^2\right) -ikU_{yy}\right]\phi \\
&+D\left(2R \Ndg S U D \phi\right),
\label{Orr-somerfield}
\end{split}
\end{equation}
where $D=d/dy$, and subject to the boundary conditions $D\phi = D^2\phi = 0$ at $y=0$ and $y=2$.

%We solve this eigenvalue problem 
%and in fig\ref{Re_vs_delta} we plot the frequency (Im$(\sigma)$) of the fastest growing mode with that of experimental observations of frequencies in
%the lab scale experiments, for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field. 
%The observed frequencies are associated with the peaks in velocity spectra, frequency of $\monami$ and frequency of 
%vortex passage \cite{Ghisal02}, and compare well with the predicted frequencies.

We solve this eigenvalue problem and a threshold in $\Rey$ of O(100), above which the flow is unstable (Re($\sigma$)$>$0), emerges from the 
solution of \eqref{Orr-somerfield}. The dependence of this critical $\Rey$, and the corresponding fastest growing wavenumber $k$, on $\delta/H$ and $\hg/H$ is shown 
in figure \ref{Re_vs_delta}, and is found to compare well with experimental observations\cite{Ghisal02}.
The threshold Reynolds number increases with the grass density, indicating a competition between the 
destabilizing shear in the flow, and the stabilizing effect of damping due to vegetation drag.
Previous linear stability calculations for terrestrial grass do not exhibit a threshold flow condition 
either due to the exclusion of the vegetation drag in their model\cite{Raupach96}, or because of ad-hoc
assumptions such as for the mean velocity profile\cite{Raupach96,Delangre06}.

Comparison of frequency (Im$(\sigma)$) of the fastest growing mode with that of experimental observations of frequencies in
the lab scale experiments, for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field, is shown in the inset in figure \ref{Re_vs_delta}b. 
The observed frequencies are associated with the peaks in velocity spectra, frequency of $\monami$ and frequency of 
vortex passage \cite{Ghisal02}, and compare well with the predicted frequencies. 
Regardless of the assumptions, the introduction of vegetation drag is observed to damp the linear modes\cite{Delangre06}.
